---
title: "Working With Data 1"
authors: Andrea Paz-Lacavex, Lance Takata, Paige Gardner
format: html
editor: visual
---

## 

In our Seaside Chat, we will be teaching about the use of some of the useful functions in tidyverse to work with data, such as:

-   Working with pre-existing data sets

-   Selecting ranges of data that we're interested in exploring

-   Practice with data manipulation

*This activity assumes students are somewhat familiar with the tidyverse package*

# Learning Outcomes

By the end of our Seaside Chat, students will be able to:

-   Use tidyverse functions mutate(), case_when() to successfully create new columns in your dataframe

## Why did we pick this topic?

As you learn new methods for working with data in R, it is important to practice those skills. For more experienced users, it is just as important to continue learning new skills that may improve your effectiveness and efficiency.

We will be working with a pre-existing dataset about crabs. You will need to load the package and the data into R.

Let's get into it!

## 1. Set Up

Load in the required packages and data. Open the data frame and notice the column names. We will have you run some basic functions to check your understanding of the data and structure.

```{r}
#Run this line if you do not have tidyverse already installed
install.packages("tidyverse")
library(tidyverse)

# Load in the package and load in the data
install.packages("HDclassif")
library("HDclassif")

#Load data
data(Crabs)

#Let's explore the data!
#First, let's check the dataframe by clicking on the top-right environment and click on Crabs. You'll see that some of the column names are not very descriptive. Run the followling line to check what the column names represent:

?Crabs

```

You'll see the class column contains two values: the species and the sex. Let's separate them so we can work with both species and sex as individual columns!

## 2. Create new columns using mutate and basic statistical functions

As said before, we want two new columns: species and sex. Enter: mutate!

Before we do that, let's practice a bit more with mutate and what it does.

**What is mutate?** The mutate function is used to both modify existing column in a dataframe or create new ones without messing with any of the other columns. Cool! That means we can do things like add columns together, add an average column, pull out the year from a date, and much more.

Let's start with some basics!

```{r}
# Question: What is the area of a crab's carapace? Create a new column called carapace_area for each individual (row)
Crabs <- Crabs %>% 
  mutate(carapace_area = CL*CW)

```

We can also temporarily mutate columns without permanently committing them to the data frame. *Reminder: By putting Crabs\<- up above, I was creating permanent changes to the data frame. This can be undone by just reloading the data!*

Let's temporarily change the data frame. All of the measurements are in mm. Let's make it into cm!

```{r}
Crabs %>% 
  mutate(CL = CL*10,
         CW = CW*10,
         BD = BD*10,
         FL = FL*10,
         RW = RW*10)
```

Why would you want to do this? You can "pipe in" or %\>% these mutate functions into other things. For example, maybe you want to use ggplot to plot just the frontal lobe size in cm instead of mm. How might you do that?

```{r}
# We started this exercise for you, but you should go ahead and put your ggplot skills to use! Hint: notice for the data mapping in ggplot, when you pipe in and manipulate the data frame, you use a "." where you write data = .
Crabs %>% 
  mutate(FL = FL*10) %>% 
  ggplot(data = ., )
```

Now let's try a cool function! The class column currently lists the species and sex. We will separate it into two separate columns.

**Question: What do you think the str_sub function means? Go ahead and google it and put your answer below. Leave a descriptive comment about what each part of the line is doing.**

```{r}

Crabs <- Crabs %>% 
  mutate(
    Species = str_sub(class, 1, 1),
    Sex = str_sub(class, 2, 2)
  ) 
Crabs
```

Another way you can achieve the same result:

```{r}
Crabs2 <- Crabs %>%
  separate(
    col = class,
    into = c("species", "sex"),
    sep = 1,
    remove = FALSE
  ) %>% # This second part is optional but helpful if you forget what the letters mean!
  mutate(
    species = case_when(
      species == "B" ~ "Blue",
      species == "O" ~ "Orange",
      TRUE ~ species
    ),
    sex = case_when(
      sex == "M" ~ "Male",
      sex == "F" ~ "Female",
      TRUE ~ sex
    )
  )
Crabs2

```

## 3. Plot

Now, let's see do a quick plot and use Crabs2 to see if there are any differences by species by sex.

```{r}
ggplot(Crabs2, aes(x = CL, fill = sex)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  facet_wrap(~ species) +
  labs(
    x = "Carapace length (mm) (CL)",
    y = "Count",
    fill = "Sex",
    title = "Crab size (CL) by species and sex"
  ) +
  theme_minimal()

```

### Question 3.1: What differences can you see?

## 4. Applications

We are interested in using this data to communicate the number of crabs that are of harvestable length. However, we are uncertain whether carapace length (CL) or carapace width (CW) will provide a more conservative estimate. Let's start by working with carapace length (CL) first.

We will use the column "CL Carapace Length (mm)" along with the \`case_when()\` function to create a new column named "CL_harvestable_state." In this new column, we will categorize crabs as follows: those with a length in the range of 30 to 40 mm will be labeled as "CL_Harvestable," anything under 29.9 mm will be labeled as "CL_Non-harvestable," and anything over 40.1 mm will be categorized as "CL_Non-harvestable_reproductive."

```{r}
# Update our new dataframe (Crabs3) with the desired columns
Crabs3 <- Crabs2 %>%
  mutate(
    CL_harvestable_state = case_when(
      CL >= 30 & CL <= 40 ~ "CL_Harvestable",
      CL <= 29.9 ~ "CL_Non-harvestable",
      CL >= 40.1 ~ "CL_Non-harvestable_reproductive",
      TRUE ~ NA_character_
    )
  )
Crabs3
```

### Question 4.1

For the same df, please copy and paste the code above, but modify it for the "CW Carapace Width (mm)" data. For this new column, named "CW_harvestable_state," use the following ranges: categorize lengths from 34 to 44 mm as "CW_Harvestable," anything under 33.9 mm as "CW_Non-harvestable," and anything over 44.1 mm as "CW_Non-harvestable_reproductive."

```{r}

```

Now, let's use count() to count the amount of CL_harvestable and CW_harvestable:

```{r}
CL_counts<- Crabs3 %>% 
  count(CL_harvestable_state)


CW_counts <- Crabs3 %>% 
   count(CW_harvestable_state) 
```

### Question 4.3
Now, use what you've learned to make and respond to an additional questions such as: what measurement preserves the largest amount of reproductive females?

